---
title: "R Tutorial"
author: "Andie Creel"
date: "January, 2023"
output: pdf_document
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())
```
Special thanks to Sam Maher who wrote an R Tutorial in 2018 that heavily influenced this.  


# Goal
This is R Studio. The goal of this lecture is to see some basics about R so that we can dive into more exciting things tomorrow. 

## Overview of R Studio 
Layout of RStudio: 

- Script
  - Where you will be writing your own programs
- Environment/../Git
  - Mostly just environment and Git
  - Show which data objects you have loaded in memory
  - Eventually where you'll do version control with GitHub (last day)
- Files/Plots/Packages/Help/Viewer  
  - Helps you load packages and other files to load
  - Default window when you're trying to get with a function from a package (we will get to this )
- Console/Terminal
  - where the code actually run 
  - Scripts executes in console 
  - Code disappear in the console, whereas a script saves your code
  - To run something in the console, type it in and hit "enter"
  - Important to know there is a "terminal" in R studio. Again, just know it's there. 
  

## Console
We can type code straight into the console and run it. The console in R Studio knows your running R code. 


```{r}
# Type the following in the console, then press enter 
2+3


# variable assignment happens with an arrow (on a mac can do option + -)
a <- 2
b <- 3

a + b

```


## How to write and execute a script

I wrote this pdf using an R Markdown (**show them quickly**). I love these because I'm able to write lots of notes to myself while I'm coding, and then it produces a nice shareable file where I can share my notes, code and results with others. You can do the minipsets with a script or r markdown.

If you have no experience with R, start with a script. If you want to do a R Markdown and am used to them, that's fine. 

- New Script: File > New File > R Script (or R Markdown, which is what I'm using now), or just the New Document script in the upper left hand corner of the screen > R Script (or R Markdown)
-	Keyboard shortcut to run a section of code: highlight or put your cursor on that line and hit Ctrl + Enter (Windows) or Command + Enter (Mac)
- Executing code: the run button at the top right corner of the script 

For R Markdowns, we also have the Knit button at the top of the script. That will run all code and "knit" it together into a pdf or HTML or doc (which is specified at the top of the R Markdown file)


You can "clean up" the Environment after you've executed code by clicking the broom icon. **This will delete everything in your environment**.

## Basic Data Types

```{r}
# this is a comment, you can use '#' to write notes to yourself in your code 

# Numeric -- integer: no decimal points
myInt <- 1

# Numeric -- double: decimal points
myNum <- 2.4

# logical (Boolean): a true/false statement
myBool_1 <- (3 < 4)
myBool_2 <- (3 > 4)

# character (string) 
myChar_a <- "a"
myChar_b <- 'b'

```

## Ways to store datatypes 

```{r, results='hide'}
# vector: can only be a vector of one data type (numeric, logical, string)
myVec_n <- c(1, 2, 3, 4, 5)
myVec_s <- c(str, "b", "c")

# matrix: can only be a matrix of one data type
myMat_n <- matrix(c(myVec_n, 
                6, 7, 8, 9, 10), 
              nrow = 2, 
              ncol = 5)

# Lists: Very powerful, but somewhat confusing. For now, just know they exist
myList <- list(2, "c", myMat_n)
myList[[1]] # returns numeric 
myList[[2]] # returns string
myList [[3]] # returns matrix

# data frame: can have multiple data types  
myDF <- as.data.frame(myMat_n)
colnames(myDF) # these don't mean anything to me 
colnames(myDF) <- c("age_yr", "weight_lb", "income_$", "hieght_ft", "height_in")

```

Dataframes are like matrices, but we can have different data types in each column. We can also reference specific columns using the “$” operator, followed by the name of the column we want to look at.

```{r, results='hide'}
# investigate one column 
myDF$age_yr

#create a new column 
myDF$nonsense <- myDF$age_yr + myDF$weight_lb

```

For the most part, you’ll be loading new data in as tables, but you might have to create one at some point. By looking at how they’re created, we can get a better sense of what goes into them.

```{r, results='hide'}
# Create the data frame
BMI <- 	data.frame(
   gender = c("Male", "non-binary","Female"), 
   male = c(T, F, F),
   height = c(152, 171.5, 165), 
   weight = c(81, 93, 78),
   Age = c(42,38,26)
)

# Try referencing one column
BMI$male # version 1
BMI[,2] #version 2

# Try referencing one row 
BMI[1,]

# Try referencing one cell
BMI$height[1] # version 1
BMI[1,3] # version 2


```


## Functions, loops, and if statement

Functions: once you have initialized them, they take in a set of data, perform a set of operations on them, and then give you some return value

```{r, results='hide'}

myF <- function(x){
  y <- x - x^2
  return(y)
}

myF(.5)
myF(.25)
myF(.7)

```

for loops: iterates through a task for a set number of times 

```{r, results='hide'}

#simple 
for (i in 1:4){
  print(i)
}

# combining loop and function  
for (i in 1:4){
  
  y = myF(i/4)
  
  print(y)
}


#manipulating a data frame (follows version one of referencing one cell above)
for (i in 1:length(BMI$Age)) {
  BMI$Age[i] <- i
}
```

If statements: sometimes you want to execute a task ONLY if a certain condition is met. 

```{r, results='hide'}
# goes through each row and changes age if someone is male
for (i in 1:length(BMI$male)) {
 
   if (BMI$male[i] == TRUE) {
    BMI$Age_new[i] <- BMI$Age[i] - 3
  }else{
    BMI$Age_new[i] <- BMI$Age[i]
  }

}

```

## Other R Tutorials

[Princeton Getting Started with R](https://dss.princeton.edu/training/)

[UCLA Getting Started with R](https://stats.oarc.ucla.edu/stat/data/intro_r/intro_r_interactive_flat.html)


## Some specific packages
We haven't covered packages yet, but a few good resources for the tomorrow.

[ggplot](http://zevross.com/blog/2014/08/04/beautiful-plotting-in-r-a-ggplot2-cheatsheet-3/)

[dplyr and tidyr](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)

<!-- [R and GIS](https://cran.r-project.org/web/views/Spatial.html) -->

<!-- [R Shiny (interactive plots)](https://shiny.rstudio.com/gallery/) -->

