---
title: "Day 1 Mini Problem Set for R Tutorial"
author: "Your Name"
date: "Date"
output: html_document
---

# Introduction

Try to solve each problem using day one's material and some minimal Googling. If you get stuck, please come to office hours which are 2-5pm in 301 prospect, room 101. 

Remember to consider writing pseudo code prior to trying to code your answer to the problem! 

This problem set should take 1-2 hours. If you feel like you're struggling through getting things to click, don't get discouraged. Struggling through a problem is the best way to learn it! (If it starts to get discouraging, come to office hours!)


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Problem 1: Basic Operations

Create two numeric variables with any values and perform basic arithmetic operations (addition, subtraction, multiplication, division) on them.
Assign the result of a division to a new variable and print it.

```{r}
# Insert code here

```


# Problem 2: Working with Vectors
Create a numeric vector with at least 5 elements.
Compute the sum (`sum()`) and mean (`mean()`) of the vector.
Use a logical comparison (true/false) within a for-loop to identify which elements of the vector are greater than the mean. You may need to use the `print()` command.

```{r}
# Insert code here

```


# Problem 3: Basic Data Frame Manipulation
Create a data frame with at least 3 columns and 5 rows. The columns should include a mix of numeric, character, and logical data types.
Print the entire data frame.
Print only the second column of the data frame.

```{r}
# Insert code here

```

# Problem 4: Loop and Function
Write a function that takes a numeric input and returns the square of the number.
Use a for loop to apply this function to each element of a numeric vector that you create.
Store the results in a new vector, `newVec <- rep(NA, 5)`, and print it using the `print()` command.

```{r}
# Insert code here

```

# Problem 5: Conditional Logic
Using the data frame created in Problem 3, write an if statement inside a loop to perform a conditional operation (e.g., if a number in a numeric column is even, replace it with half its value).
Print the modified data frame.

Practicing Googling to figure out how to code if a number is even or not.

```{r}
# Insert code here

```


# Problem 6: Reflect
Write a brief reflection (2-3 sentences) on what you found challenging and what you learned from this problem set.

<!-- Write your reflection here -->