---
title: "Day 1, Lecture 2: Base R"
author: "Andie Creel"
date: "January, 2023"
output: 
  pdf_document:
    number_sections: true
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())
```
Special thanks to Sam Maher who wrote an R Tutorial in 2018 that influenced this.  


# Goal
This is R Studio. The goal of this lecture is to see some basics about R so that we can dive into more exciting things tomorrow. 

# Overview of R Studio 
Layout of RStudio: 

- Script
  - Where you will be writing your own programs
- Environment/../Git
  - Mostly just environment and Git
  - Show which data objects you have loaded in memory
  - Eventually where you'll do version control with GitHub (last day)
- Files/Plots/Packages/Help/Viewer  
  - Helps you load packages and other files to load
  - Default window when you're trying to get with a function from a package (we will get to this )
- Console/Terminal
  - where the code actually run 
  - Scripts executes in console 
  - Code disappear in the console, whereas a script saves your code
  - To run something in the console, type it in and hit "enter"
  - Important to know there is a "terminal" in R studio. Again, just know it's there. 
  

# Console

- We can type code straight into the console and run it. 
- The console in R Studio knows your running R code. 


```{r}
# Type the following in the console, then press enter 
2+3


# variable assignment happens with an arrow (on a mac can do option + -)
a <- 2
b <- 3

a + b

```


# How to write and execute a script

**Writing Scripts and Running Code**

- New Script: File > New File > R Script (or R Markdown, which is what I'm using now), or just the New Document script in the upper left hand corner of the screen > R Script (or R Markdown)
-	Keyboard shortcut to run a section of code: highlight or put your cursor on that line and hit Ctrl + Enter (Windows) or Command + Enter (Mac)
- Executing code: the run button at the top right corner of the script 
- For R Markdowns, we also have the Knit button at the top of the script.
  - That will run all code and "knit" it together into a pdf or HTML or doc
  - File type is specified at the top of the R Markdown file

**R Markdown**

- I wrote this pdf using an R Markdown (**show them quickly**). 
- I love these because I'm able to write lots of notes to myself while I'm coding
  - it produces a nice shareable file 
  - can share my notes, code and results with others. 
- You can do the minipsets with a script or r markdown.

**Mini psets**

- If you have no experience with R, start with a script. 
- If you want to do a R Markdown and are used to them, that's fine. 

You can "clean up" the Environment after you've executed code by clicking the broom icon. **This will delete everything in your environment**.

# Basic Data Types and COMMENTS

```{r}
# This is a comment, you can use '#' to write notes to yourself in your code 
# - Comments are what make or break good coders, and coders who can collaborate with others. 
# - If you ever thing you writing "too" many comments, you almost always not.
# - The things you think are obvious in your code wont be to others 
# - (nor yourself in a year when you get back to a project)

# Numeric -- integer: no decimal points
myInt <- 1

# Numeric -- double: decimal points
myNum <- 2.4

# logical (Boolean): a true/false statement. Use () to evaluate if something is true or false
myBool_1 <- (3 < 4)
myBool_2 <- (3 > 4)

# character (string) 
myChar_a <- "a"
myChar_b <- 'b'

```

# Ways to store datatypes 

```{r}
# vector: can only be a vector of one data type (numeric, logical, string)
myVec_n <- c(1, 2, 3, 4, 5)
myVec_s <- c(myChar_a, "b", "c")

# matrix: should only be a matrix of one data type
myMat_n <- matrix(c(myVec_n, 
                6, 7, 8, 9, 10), 
              nrow = 2, 
              ncol = 5)

# Lists: Very powerful, but somewhat confusing. For now, just know they exist
myList <- list(2, "c", myMat_n)
myList[[1]] # returns numeric 
myList[[2]] # returns string
myList [[3]] # returns matrix
```

## Dataframes

- Like matrices
- Can have different data types in each column
- Reference specific columns using the “$” operator, followed by the name of the column
- For the most part, you’ll be loading new data by reading a csv 
- You might have to create one at some point.
- By looking at how they’re created we can get a better sense of what goes into them

```{r}
# data frame: can have multiple data types 
myDF <- as.data.frame(myMat_n)
colnames(myDF) # these don't mean anything to me 
colnames(myDF) <- c("age_yr", "weight_lb", "income_$", "height_ft", "height_in")

```


```{r}
# investigate one column 
myDF$age_yr

#create a new column 
myDF$nonsense <- myDF$age_yr + myDF$weight_lb

```



```{r}
# Create the data frame
myPpl <- 	data.frame(
   gender = c("Male", "non-binary","Female"), 
   male = c(T, F, F),
   height = c(152, 171.5, 165), 
   weight = c(81, 93, 78),
   age = c(42,38,26)
)

# Try referencing one column
myPpl$male # version 1
myPpl[,2] #version 2

# Try referencing one row 
myPpl[1,]

# Try referencing one cell
myPpl$height[1] # version 1
myPpl[1,3] # version 2


```


# A word of caution 

- Make sure you don't over write your variables by accident. 

```{r}

# assigning new value to same variable (something to do carefully)
a <- 5
a <- a + 1 # If you run this line more than one, you will NOT get six
a

# assigning new value to new variable
a <- 5
a_new <- a + 1 # If you run this line more than one, you WILL get six
a_new
```


# Functions 

Functions: once you have initialized them, they take in an input, perform a set of operations on them, and then give you some return value. 

**Example on board**

- consider the function: myF(x) { y <- x + 3; return(y)}
- what does myF(3) return? 6 

Points:

- These are helpful when you have something that you do often 
- Rule of thumb: if you're copying and pasting code 3 times or more, make function 
- (i say if you are going to copy past ever, because even if you think it'll only be twice it'll probably be more)
- Recent example for me:
  - wrote a function to take a date and return the season
  - Wrote a function to get kelvin and return Fahrenheit 


```{r}

myF <- function(x){
  y <- x - x^2
  return(y)
}

myF(.5)
myF(.25)
myF(.7)

```


# Loops

- for loops: iterates through a task for a set number of times 
- Consider these loops (psuedo code): 
  - For (i in 1 through 4) { print i }
  - For (i in 1 through 4) { print i / 4}
- Can be helpful when 
  - Iterate through a column of data and do something to each row 
  - Construct a new column and want to construct each row by scratch


```{r}

#simple 
for (i in 1:4){
  print(i)
}

# more involved 
for (i in 1:4){
  print(i/4)
}

# combining loop and function  
for (i in 1:4){
  
  y = myF(i/4)
  
  print(y)
}


# manipulating a column (from above: version one of referencing a cell)
for (i in 1:length(myPpl$age)) {
  myPpl$age[i] <- myPpl$age[i] + 1 # everyone aged one year
}

# Making a new column 
for (i in 1:length(myPpl$age)) {
  myPpl$age_new[i] <- myPpl$age[i] + 1 # everyone aged one year
}
```


# If statements

- sometimes you want to execute a task ONLY if a certain condition is met. 
- Open the myPpl df:
  - Our RA did not record men's ages right 
  - All men are actually 3 years younger than what's recorded
  - What would the correct DF look like?
- If statements let you fix a mistake like this
- Also demonstrates why the Boolean (true/false or indicator) variable is so powerful

```{r}
# goes through each row and changes age if someone is male
for (i in 1:length(myPpl$male)) {
 
   if (myPpl$male[i] == TRUE) {
    myPpl$age_new_m[i] <- myPpl$age[i] - 3
  }else{
    myPpl$age_new_m[i] <- myPpl$age[i]
  }
}

```


# Other R Tutorials

[Princeton Getting Started with R](https://dss.princeton.edu/training/)

[UCLA Getting Started with R](https://stats.oarc.ucla.edu/stat/data/intro_r/intro_r_interactive_flat.html)

[git lab intro](https://rgup.gitlab.io/research_cycle/01_intro.html#class_notes)

# Some specific packages
We haven't covered packages yet, but a few good resources for the tomorrow.

[ggplot](http://zevross.com/blog/2014/08/04/beautiful-plotting-in-r-a-ggplot2-cheatsheet-3/)

[dplyr and tidyr](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)

<!-- [R and GIS](https://cran.r-project.org/web/views/Spatial.html) -->

<!-- [R Shiny (interactive plots)](https://shiny.rstudio.com/gallery/) -->

